Staged Abstract Interpreters

### 2019-09-04 paper/50_stageabs.tex
Proofread Section 7

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters from an \textit{operational} perspective.
In this section, we review and summarize our recipe to achieve the staged
abstract interpreter, discuss correctness issues, and different design choices.

@@

\begin{itemize}
  \item Constructing a generic interpreter that abstracts over binding-time,
    value domains, and primitive operations. In this paper, the generic interpreter
    is implemented in monadic style; therefore, the semantics can be encapsulated
    into monads.
  \item By using an appropriate monad stack, we can implement a modular, unstaged,
    abstract interpreter. This step has been explored in the previous literature.
  \item Replacing the monad stack to staged monad stack, such monads operate on
    staged data types, i.e., next-stage values. We also need to reimplement several
    primitive operations on staged values, in a modular fashion.
\end{itemize}

Monadic interpreters are known to be able to decouple the interpretation process
and the underlying semantics. The key insight in this paper is that by making the
monadic interpreter binding-time polymorphic, it can be extended for generating
efficient code in a staging framework. The underlying semantics and staging are
two orthogonal dimensions. It is important to note that the computation
encapsulated by the monads are not staged; only data (such as sets and maps) are
staged. All the monadic computation, i.e., functions passed to the monadic bind
operation @flatMap@, are statically known. This is why we can eliminate the
monadic layer and its associated overhead.

@@

\paragraph{Monadic-style vs Direct-style} We use monadic interpreter through the paper,
but it is not necessary for staging. One can inline the monadic
operations and obtain an abstract interpreter in continuation-passing
style, or even translate back to direct-style that may use explicit
side-effects such as mutations. In either case, we can still apply the
staging idea to the abstract interpreter and remove the interpretation
overhead. Although, monads make the staged abstract interpreter can be
implemented in a modular and extensible way.

>>>>

We have gradually presented the confluence of specialization and abstraction of
concrete interpreters from an \textit{operational} perspective.
In this section, we review and summarize our recipe to achieve the staged
abstract interpreter and discuss correctness issues and different design choices.

@@

\begin{itemize}
  \item Constructing a generic interpreter that abstracts over binding times,
    value domains, and primitive operations. In this paper, the generic interpreter
    is implemented in monadic style; therefore, the semantics can be encapsulated
    into monads.
  \item By using an appropriate monad stack, we can implement a modular, unstaged,
    abstract interpreter. This step has been explored in the previous literature.
  \item Replacing the monad stack to a staged monad stack; such monads operate on
    staged data types, i.e., next-stage values. We also need to reimplement several
    primitive operations on staged values, in a modular fashion.
\end{itemize}

Monadic interpreters are known to be able to decouple the interpretation process
and the underlying semantics. The key insight in this paper is that by making the
monadic interpreter binding-time polymorphic, it can be extended for generating
efficient code in a staging framework. The underlying semantics and staging are
two orthogonal dimensions. It is important to note that the computation
encapsulated by the monads are not staged: only data (such as sets and maps) are
staged. All the monadic computation, i.e., functions passed to the monadic bind
operation @flatMap@, are statically known. This is why we can eliminate the
monadic layer and its associated overhead.

@@

\paragraph{Monadic-style vs Direct-style} We use a monadic interpreter throughout the paper,
but it is not necessary for staging. One can inline the monadic
operations and obtain an abstract interpreter in continuation-passing
style, or even translate back to a direct-style that may use explicit
side-effects such as mutations. In either case, we can still apply the
staging idea to the abstract interpreter and remove the interpretation
overhead. However, monads allow the staged abstract interpreter to be
implemented in a modular and extensible way.
